---
title: "Proj3 Workspace Em"
author: "Em Webster"
date: "4/29/2018"
output: html_document
---
** load all packages
```{r}
library(dbplyr)
library(mdsr)
library(RMySQL)
library(readr)
library(ggthemes)
```
```{r}
db <- dbConnect_scidb("citibike")
```

** show all tables in citibike
```{sql connection=db}
SHOW TABLES;
```

** explore station_months
```{sql connection=db}
DESCRIBE station_months
```

** explore station_summary
```{sql connection=db}
DESCRIBE station_summary
```

** explore trip_summary
```{sql connection=db}
DESCRIBE trip_summary
```

** explore trips
```{sql connection=db}
DESCRIBE trips
```

** deep explore trips
```{sql connection=db}
SELECT * FROM trips 
WHERE start_time LIKE '2017-03-29%'
LIMIT 0,10
```

** deep explore station_summary
#```{sql connection=db}
#SELECT * FROM station_summary LIMIT 0,10
#```

** make gender pct from trips
#```{sql connection=db}
SELECT
user_type,
sum(if(gender = 1, 1, 0)) / sum(1) AS male_pct,
sum(if(gender = 2, 1, 0)) / sum(1) AS female_pct
FROM trips AS t
WHERE start_time LIKE '2017-03-29%'
LIMIT 0,5
```

```{sql connection=db}
SELECT
user_type,
num(*) AS male_pct,
FROM trips AS t
WHERE start_time LIKE '2017-03-29%'
LIMIT 0,5
```

#```{r}
genderpct <- "
SELECT 
user_type,
sum(if(gender = 1, 1, 0)) / sum(1) AS male_pct,
sum(if(gender = 2, 1, 0)) / sum(1) AS female_pct
FROM trips AS t
LIMIT 0,5
"
db %>% dbGetQuery(genderpct) %>% collect()
genderpct <- as.data.frame(genderpct)
genderpct
```

```{r}
query <- "
SELECT 
user_type,
gender
FROM trips AS t
WHERE gender = 1 OR gender = 2
LIMIT 0,100
"
query <- db %>% dbGetQuery(query) %>% collect()
genderpct <- as.data.frame(query)

```

```{r}
genderpct
```


```{r}
genderpct$gender <- as.factor(genderpct$gender)
pie <- ggplot(genderpct, aes(x = "", fill = factor(gender))) + 
  geom_bar(width = 1, color='black') +
  theme(axis.line = element_blank(), 
        plot.title = element_text(hjust=0.5)) + 
  labs(fill="gender", 
       x=NULL, 
       y=NULL, 
       title="Pie Chart", 
       caption="Source: citi")
pie <- pie + coord_polar(theta = "y", start=0) + scale_fill_manual(values=c("dodgerblue", "deeppink"), 
                       name="Gender",
                       breaks=c("1", "2"),
                       labels=c("Male", "Female")) + theme_fivethirtyeight()

pie
```

```{r}
var <- genderpct$gender
nrows <- 10
df <- expand.grid(y = 1:nrows, x = 1:nrows)
categ_table <- round(table(var) * ((nrows*nrows)/(length(var))))
categ_table
```

<<<<<<< HEAD
**
```{sql connection=db}
SELECT 
sum(duration)/sum(1) AS zero_duration
FROM trips
WHERE gender = '0'
LIMIT 0, 10;
```


```{sql connection=db}
SELECT 
sum(if(duration > 0, duration, if(gender = 1,1,0)))/sum(1) AS one_duration,
sum(if(duration > 0, duration, if(gender = 0,1,0)))/sum(1) AS zero_duration
FROM trips
LIMIT 0, 10;
```

```{sql connection=db}
SELECT 
sum(duration)/sum(1) AS two_duration
FROM trips
WHERE gender = '2'
LIMIT 0, 10;
```

```{r}
library(waffle)
font_import()
fonts()[grep("Awesome", fonts())]
```

```{r}
fonts()[grep("Awesome", fonts())]
loadfonts()
```

```{r}
library(extrafont)
fonts()[grep("Awesome", fonts())]
```

```{r}
library(waffle)
parts <- c(`Male (73%)`=73, `Female (27%)`=27)
chart <- waffle(parts, rows=5, use_glyph = "bicycle", glyph_size = 6,colors = c("dodgerblue", "deeppink"),
       title="Citi Bike Users by Gender") + theme_fivethirtyeight() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),axis.text = element_blank(),
        axis.ticks.x = element_blank())+ labs(subtitle="1 square = 425 Riders")
chart
```


```{r}
query <- "SELECT start_time, stop_time, gender FROM trips LIMIT 0,300"

trip_times <- db %>% dbGetQuery(query) %>% collect()
trip_times
```

```{r}
trip_times <- trip_times %>% filter(gender != 2) %>% mutate(trip_length = difftime(stop_time, start_time, units = "mins"), row_num = row_number()) 
trip_times <- trip_times %>% mutate(gender = as.character(gender))
trip_times
```

```{r}
trips_plot <- trip_times %>% ggplot(aes(x = gender, y = trip_length, fill = gender )) + geom_boxplot() + ylab("Trip Duration in Minutes") + xlab(" ") + ylim(c(0,65)) + scale_fill_manual(name="Gender",
                         breaks=c("0", "1"),
                         labels=c("Male", "Female"),
                         values = c("dodgerblue", "deeppink")) + theme_fivethirtyeight() +theme(axis.title=element_text(size=12)) + ggtitle("Trip Duration by Gender") +
  stat_summary(geom="text", fun.y=median,
               aes(label=sprintf("%1.1f", ..y..)),
               position=position_nudge(x=0.50), size=3.5)

trips_plot
```

**Age by Trip duration
#```{r}

query<-"SELECT birth_year, gender, user_type,start_time, stop_time FROM trips
WHERE start_time LIMIT 0,100"

query<-db %>% dbGetQuery(query) %>% collect()

age_time<-query %>% mutate(age=2017-birth_year, row_num=row_number(),trip_length = difftime(stop_time, start_time, units = "mins"), men=gender==1, women=gender==2, gender = as.character(gender))%>% filter(age !=2017, gender !=0) %>% filter(birth_year !=0,age!= 131)
##Can't get the right color or description for male and female labels 1 should be male 2 should be female
age_time_plot<-age_time%>%ggplot(aes(x=trip_length, y=age, color=gender))+geom_point()+geom_smooth()+ xlab("Trip Duration in Minutes")+ ylab("Age in 2017")+scale_fill_manual(name="Gender",
                         breaks=c("1", "2"),
                         labels=c("Male", "Female"),
                         values = c("dodgerblue", "deeppink")) + theme_fivethirtyeight() +theme(axis.title=element_text(size=12)) + ggtitle("Trip Duration by Age and Gender")

age_time_plot


#```

**Age by gender box plot 
```{r}
query<-"SELECT birth_year, gender, user_type,start_time, stop_time FROM trips
WHERE start_time LIMIT 0,100"

query<-db %>% dbGetQuery(query) %>% collect()

age_time<-query %>% mutate(age=2017-birth_year, row_num=row_number(),trip_length = difftime(stop_time, start_time, units = "mins"), men=gender==1, women=gender==2, gender = as.character(gender))%>% filter(age !=2017, gender !=0) %>% filter(birth_year !=0,age!= 131)

##same labeling issues
age_time_plot <- age_time %>%ggplot(aes(x= gender, y=age, fill=gender))+geom_boxplot()+xlab(" ")+ylab("Age")+ scale_fill_manual(name="Gender",
                         breaks=c("1", "2"),
                         labels=c("Male", "Female"),
                         values = c("dodgerblue", "deeppink"))+theme_fivethirtyeight() +theme(axis.title=element_text(size=12)) + ggtitle("Age by Gender") + scale_y_continuous(breaks = c(0,10,20,25,30, 35, 40, 45, 50, 55, 60 , 65, 70 , 75))

age_time_plot


```







