---
title: "This Will Be the Final Writeup!"
author: "NAMES"
date: "05/19/2018"
output:
  html_document: default
  pdf_document:
    number_sections: yes
urlcolor: blue
---


******************
IGNORE ALL THE OTHER STUFF BELOW RIGHT NOW, THIS IS THE OLD MARKUP FILE FROM MY PROJ ONE AND IM GONNA USE IT TO MODEL OUR WRITEUP
******************


```{r include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE, cache=FALSE}
knitr::opts_chunk$set(error = TRUE)
```

```{r include=FALSE, cache=FALSE}
library(tidyverse)
library(readr)
library(ggthemes)
```

```{r include=FALSE, cache=FALSE}
tuition_wage <- read_csv("average-tuition-wage.csv")
```

```{r include=FALSE, cache=FALSE}
cost_plot <- ggplot(tuition_wage, aes(x = year))
```

```{r include=FALSE, cache=FALSE}
cost_plot <- cost_plot +
  geom_bar(
    stat = "identity",
    aes(y = summer_pay),
     fill = "darkseagreen3", color = "white")
```

```{r include=FALSE, cache=FALSE}
cost_plot <- 
  cost_plot +
  geom_line(aes(y = public_4yr_instate),
    color = "firebrick2")
```

```{r include=FALSE, cache=FALSE}
cost_plot <-
  cost_plot +
  geom_line(aes(y = private_4yr),
    color = "dodgerblue2")
## anyone know how to mess with this so it is more readable?
  ## This is a start but ill continue to the work to make it cute and journalism worthy 
```

```{r include=FALSE, cache=FALSE}
# Maybe no subtitle and be more specific in the headings on graph?

cost_plot <- cost_plot + labs( x = "Year", y = "Dollars") + ggtitle("Average Tuition Fees at Four Year Institutions vs. 
Minimum Wage Summer Earnings", subtitle = "Public Instate and Private Schools" ) + theme(plot.title = element_text(hjust = 0.5)) + theme(plot.subtitle = element_text(hjust = 0.5))
```


```{r include=FALSE, cache=FALSE}
cost_plot <-
  cost_plot +
  geom_bar(
    stat = "identity",
    aes(
      y = ifelse(
        year == 1988,
        public_4yr_instate,
        0
      )
    ),
     fill = "firebrick2",
    color = "white"
  )
```

```{r include=FALSE, cache=FALSE}
cost_plot <-
  cost_plot +
  geom_bar(
    stat = "identity",
    aes(
      y = ifelse(
        year == 1969,
        private_4yr,
        0
      )
    ),
     fill = "dodgerblue2",
    color = "white"
  )
```

```{r include=FALSE, cache=FALSE}
cost_plot <- cost_plot +
    geom_text(
    x = 1975, y = 7000, 
    label = "Private Schools",
    colour = "dodgerblue2"
    ) +
  geom_text(
    x = 1995, y = 6000, 
    label = "Public Schools in State ",
    colour = "firebrick2"
    )
```

```{r include=FALSE, cache=FALSE}
library(readr)
library(scales)
schoolname <- read_csv("specific-schools-edited.csv")
```

```{r, echo = FALSE}
name_plot <- ggplot(schoolname, aes(x=university, y = (farm_min_wage_1976*480)/cost_1976)) + geom_bar(stat="identity", position="dodge", aes(fill=factor(Year))) + scale_y_continuous(labels=percent_format()) + xlab("University") + ylab("Percetage of University Fees")
name_plot <- name_plot + scale_fill_manual(values = c("firebrick1", "firebrick")) + guides(fill=guide_legend(title="Year"))
```

```{r include=FALSE, cache=FALSE}
name_plot <- name_plot + ggtitle("Percentage of University Fees that Summer Wage Covers") + theme(plot.title = element_text(hjust = 0.5)) + theme(plot.subtitle = element_text(hjust = 0.5)) 
```
## Introduction
Many politicians often turn back time to the "golden days" when they were able to pay the entirety of their college tuitions with their summer wages. We decided to put this claim into context. The 1970s marks a period in economic history that begins a general wage stagnation, which is when the cost of goods and services exceeds one's ability to work for it and afford it The cost of attending higher education is not exempt from this trend, even public institutions. Public four year institutions aim to make higher education more accessible for economically disadvantaged students, yet in recent years tuition at public institutions has risen so sharply that being able to work one's self through college has become less realistic than in the 1970s. 

## Tuition is Rising Faster than Minimum Wage
The graph above charts the increase of summer earnings over time (calculated as the the federal hourly minimum wage for 12 weeks at 40 hours a week) against the increase of public and private four year college tuition in the United States. We can see that prior to 1988, one could cover an entire year of tuition at a four year public institution with summer wages. However, due to tuition rising faster than the federal minimum wage, public higher education becomes inaccessible to those relying on minimum wage jobs around the 1990s. Private education, on the other hand, was never attainable with a summer's wage during the time frame of the data.

```{r, echo = FALSE}
library(ggthemes)
cost_plot <- cost_plot + theme_gdocs()
cost_plot
```

## How Much College Can One Pay For With Minimum Wage?
This graph depicts three well-known public universities and the percentage of university fees covered by summer wages. We focused on public universities because it lies at the nexus of the claim that public education is accessible for everyone. While this claim was true in 1976, it no longer stands in 2014. Our previous graph focused on broader trends whereas this graph hits the numbers home by focusing on three distinguished universities. 

```{r, echo = FALSE}
name_plot <- name_plot + theme_gdocs(base_size = 8)
name_plot
```

## Average Rate of Growth
This graph displays the marginal growth of Tuition and Summer Pay depicting the comparison of Tuition fees and Summer Pay marginal growth.

In the 1980's there is little to no increase in summer pay, but tuition prices continued to grow. Graph 0.1 calculated the amount of tuition that summer pay can account for. In the year 1988, summer pay stopped being able to cover a year's tuition. This, in part, is due to the previous years lacking of summer pay growth. It was also in 1988 that the summer pay began to grow. The graph does not show trends between the summer pay and tuition growth. There is, a similarity between the average tuition at both private and public institutions. The average marginal increases at all universities seemed to be similarly proportional. It was not until the year 2000 where the growth patterns differentiate. 

```{r, echo = FALSE, message=F, warning=F}
library(tidyverse)
library(readr)
plotter <- read_csv("average-tuition-wage.csv")
#view(plotter)

#For Loop to change data for Private_4yr
z = plotter[2, "private_4yr"]
for (i in 2 : 44)
{
  #Copy over years
  y = z
  x = plotter[i+1, 2]
  omega = (x/y)*100
  #print(omega)
  plotter[i-1, 2] = omega
  #Sets a variable buffer to save the value of x
  z = x
}
plotter[1,2] = 100 #Sets the original data to be 100%

#For Loop to change data for Public_4yr
z = plotter[2, "public_4yr_instate"]
for (i in 2 : 44)
{
  #Copy over years
  y = z
  x = plotter[i+1, 3]
  omega = (x/y)*100
  #print(omega)
  plotter[i-1, 3] = omega
  #Sets a variable buffer to save the value of x
  z = x
}
plotter[1,3] = 100 #Sets the original data to be 100%


#For Loop to change data for Summer_Pay
z = plotter[2, "summer_pay"]
for (i in 2 : 44)
{
  #Copy over years
  y = z
  x = plotter[i+1, 5]
  omega = (x/y)*100
  #print(omega)
  plotter[i-1, 5] = omega
  #Sets a variable buffer to save the value of x
  z = x
}
plotter[1,5] = 100 #Sets the original data to be 100%

#Initializes the graph and sets the X axis to be the year
percent_growth <- ggplot(plotter, aes(x = year)) +
  labs( x = "Year", y = "Percent Growth %") +
  ggtitle("Average Percent Growth of Tuition", subtitle = "Public Instate and Private Schools")

#Adds Slopes From Growth to Graph
percent_growth <- percent_growth +
  geom_line(aes(y = private_4yr), color = "firebrick2") +
  geom_line(aes(y = public_4yr_instate), color = "dodgerblue2") +
  geom_line(stat = "identity", aes(y = summer_pay), color = "darkseagreen3") +
  geom_text(
    x = 1975, y = 125, 
    label = "Private 4yr",
    colour = "firebrick2"
    ) +
  geom_text(
    x = 1985, y = 125, 
    label = "Public 4yr",
    colour = "dodgerblue2"
    ) +
  geom_text(
    x = 1995, y = 125,
    label = "Summer Pay",
    colour = "darkseagreen3"
  )
percent_growth <- percent_growth + theme_gdocs()
percent_growth
```

## Methodology
The data we used for this project were drawn from multiple sources. The average in-state tuition and at four-year public institutions in 1969 through 2011 came from the [National Center for Education Statistics](https://nces.ed.gov/programs/digest/d12/tables/dt12_381.asp). Data on federal minimum wages from 1939 to present came from the [Department of Labor](https://www.dol.gov/whd/minwage/chart.htm). Berkley specific data came from [here](http://admissions.berkeley.edu/cost-of-attendance), Charlottesvilles specific data came from here, and Ann Arbor specific data came from [here](http://www.umich.edu/~bhlumrec/um-fees/1976_fees.pdf) and [here](http://ro.umich.edu/tuition/tuition-fees.php).

We mapped dollars as a function of time to create space to examine, compare, and contrast our three variables as they all operate within these frames. Values were also adjusted for inflation to allow for ample comparison.
We used two graph types throughout our post: the line chart and the bar graph.

We used a line chart to illustrate our data because it accentuates trends by following a variable over time. Furthermore, a line chart helps the reader see how these trends move or don't move together and switch their order.  The information that our graph seeks to convey comes from over time, therefore we decided to use a line graph to connect our cases.The average cost of tuition against summer wages and the average rate of growth of college tuition are time based. Our decision to use line charts was also influenced by our use of averages. Because we are working with averages, we did not need to use a scatter plot examine the trend over individual cases of college tuition prices and wages to highlight the trend.

In all of our graphics wanted to choose color schemes that emphasised contrast among all the variables being measured. In our first graphic, however, we took extra precaution to ensure clarity. We came to a consensus to use bars to illustrate summer wages to keep from using too many lines and easily convey information. This also helps us focus how we aim to position the all of the trends relative to one another and gives a visual sense of what the stagnation phenomena looks like. For our second graphic, We used a stacked bar graph to create two comparisons, one comparing the cost of tuition among three individual cases (not averages) at two specific points in time. We wanted to create bold contrast between 1976 and 2014, which is why we used grey and red for the bar fill.In our third graphic, we only compared two trends, which makes it important to use  contrasting colors for clarity and a theme that did not add to line confusion. In each graphic, we selected our theme based on its ability to help focus on the data and not distract. 